---
title: "Normalization and RNA-composition"
author: "Malte Thodberg"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    smaller: true
    highlight: tango
    transition: faster
---
## Introduction

This presentations presents some background on EM-normalization for library size and RNA-composition,
as wells as some examples on how this is applied in R using the package `edgeR`.

Density curves and log-log plots will be used to explore the effects of different normalization methods.

# RNA-composition and DE

## TPM normalization
Setup simple EM:
```{r}
sample1 = c(10, 20, 30, 10, 10, 10) # Library size of 100 counts
sample2 = 2 + sample1 * 2 # Double library size
sample3 = 1 + sample1 * 3 # Triple library size
EM = data.frame(sample1, sample2, sample3)

EM
```

Note the different library sizes:
```{r}
colSums(EM)
```

## TPM normalization
TPM scaling:
```{r}
scale(EM, center=FALSE, scale=colSums(EM)) # Lets forget the M-part for now...
```

Samples can now be compared directly for analysis!

## CPM normalization
Introduce DE for some TCs
```{r}
EM.DE = EM
EM.DE[4:6,2] = EM.DE[4:6,2] * 5
EM.DE[4:6,3] = EM.DE[4:6,3] * 4

EM.DE
```
The total RNA content of sample2+3 has increased!

## CPM normalization
TPM scaling
```{r}
scale(EM.DE, center=FALSE, scale=colSums(EM.DE))
```
Non-DE genes are now under-sampled!

## CPM normalization
This can affect downstream analysis i.e. distance matrix calculations.
```{r}
dist(t(scale(EM, center=FALSE, scale=colSums(EM))))
dist(t(scale(EM.DE, center=FALSE, scale=colSums(EM.DE))))
```

# Advanced normalization

## Setup

Packages needed for the analysis:
```{r}
library(ABC2017)
library(tidyr)
library(edgeR)
library(ggplot2)
theme_set(theme_minimal()) # Make ggplots prettier
```

We will use the small `zebrafish` dataset:
```{r}
data(zebrafish)
```

The dataset is a list which contains:

- Expression: EM
- Annotation: Annotation
- Design: Information about the samples.

The same format is used for the remaining datasets in the `ABC2017` package

## Setup
EM:
```{r}
head(zebrafish$Expression)
```

Annotation:
```{r}
head(zebrafish$Design)
```

## Example data

The dataset is publically available from Bioconductor, originally comming from:

*"Silencing of odorant receptor genes by G protein betagamma signaling ensures the expression of one odorant receptor per olfactory sensory neuron"*

Ferrira *et al* 2014

Experiment on the effect of gallein treatment on Zebrafish: 

*Olfactory sensory neurons were isolated from three pairs of gallein-treated and control embryonic zebrafish pools and purified by fluorescence activated cell sorting (FACS). Each RNA sample was enriched in poly(A)+ RNA from 10–30 ng total RNA and 1 µL (1:1000 dilution) of Ambion ERCC ExFold RNA Spike-in Control Mix 1 was added to 30 ng of total RNA before mRNA isolation. cDNA libraries were prepared according to manufacturer’s protocol. The six libraries were sequenced in two multiplex runs on an Illumina HiSeq2000 sequencer, yielding approximately 50 million 100bp paired-end reads per library.*

## Plotting distributions

edgeR (via limma) provides the `plotDensities` function for exploring the effect of normalization 

```{r}
plotDensities(zebrafish$Expression, legend="topright")
```

## Plotting distributions

That did not look to good! Since the data spans multiple orders of magnitude, we can try with	 a log-scale instead.

This however brings up the problem of 0 counts - for which `log` is not defined.

The get around this probelm a small pseudo-count can be added to all counts in the EM. This does not necesarily have to be an integer, and is usually chosen to be between 0.1 and 1.0.

## Plotting distributions

```{r}
# Pseoducount and log
plotDensities(log(zebrafish$Expression+1), legend="topright")
```

## Plotting distributions

Notice how the lower quartile is zero - this means that we have a large number of genes with very low counts.

Counts with 1-3 counts are not very interesting, since they are likely to be either noise or expressed at biologically irrelevant levels. It's customary to perform som ad-hoc trimming or filtering to remove these prior to analysis.

Here we only keep genes with at least 2 counts in at least 4 samples:

```{r}
# Trim
above_one <- rowSums(zebrafish$Expression > 1)

trimmed_em <- subset(zebrafish$Expression, above_one > 3)

# Pseoducount and log
log_trimmed_em <-  log(trimmed_em + 1)
```

## Plotting distributions

```{r}
plotDensities(log_trimmed_em, legend="topright")
```

## Plotting distributions

Now we have a clearer picture of the distribution of counts within each sample. The large difference in distributions shows the need for normalization, before the samples can be compared.

As with everything in R, we do not have to recode everything from scratch. The edgeR package has a function `cpm` which has implented a large number of normalization methods and log-transformation.

edgeR does this by implementing the use of normalization factors, which is use to rescale the actual library sizes to take into account differences in RNA-composition.

## Using edgeR to normalize

Using edgeR is simple, but first we must save the EM as a `DGEList`:
```{r}
# Create DGEList-object from the trimmed em
dge <- DGEList(trimmed_em)

# Use edgeR to calculate normalization factors
dge <- calcNormFactors(object=dge, method="TMM")

# calculate log cpm values
TMM_em <- cpm(x=dge, log=TRUE, prior.count=1.0)

head(TMM_em)
```

## Using edgeR to normalize
The resulting plot shows a nicer alignment of the main peak:
```{r}
plotDensities(TMM_em)
```

## log-log plots

Another way of visualizing normalization is via a log-log plot. This is simply a scatterplot with paired expression values for two samples.

Although it only allows for pairwise comparison, it is a nice way to see the effect of normalization and the variance of expression at different levels.

## log-log plots
First we consider the (trimmed) log(counts+1)
```{r}
qplot(data=log_trimmed_em, x=Ctl3, y=Trt13, alpha=I(0.1)) + geom_smooth(method="gam") + geom_abline(color="red")
```

## log-log plots
Compare this with edgeR's TMM normalization:
```{r}
qplot(data=as.data.frame(TMM_em), x=Ctl3, y=Trt13, alpha=I(0.1)) + geom_smooth(method="gam") + geom_abline(color="red")
```

## Exercises

Team up and do the following exercises:

1. Make a plot of density curves showing the effect of all the different normalization methods implemented by `edgeR`
2. Make a plot that compares the estimated normalization factors for the `TMM` and `RLE` normalization methods.
3. Investigate the effect of different pseudocount values using log-log plots.

*HINT for 1:* Use `apply`-family and facets to compare multiple datasets. Remember to include `method="none"`.

*HINT for 2:* Read the `calcNormFactors` help file to see where the normalization factors are stored

## Question 1 code
```{r}
# Convert to a DGElist
dge <- DGEList(trimmed_em)

# Normalize using each of four methods
edgeR_methods <- c("none", "TMM", "RLE", "upperquartile")
dges <- lapply(edgeR_methods, calcNormFactors, object=dge)

# Calculate CPMs
norms <- lapply(dges, cpm, log=TRUE)
```

## Question 1 plot

```{r}
par(mfrow=c(2,2))
mapply(plotDensities, norms, edgeR_methods, MoreArgs=list(group=NULL, col=NULL, legend=FALSE))
```

## Question 2 code
```{r}
# Extract the normalization factors
norm_factors <- sapply(dges, function(x) x$samples$norm.factors)
colnames(norm_factors) <- edgeR_methods
```

## Question 2 plot
```{r}
plot(as.data.frame(norm_factors))
```

## Question 3 code
```{r}
# Create DGEList-object from the trimmed em
dge <- DGEList(trimmed_em)
dge <- calcNormFactors(object=dge, method="TMM")

# calculate log cpm values
TMM_v <- cpm(x=dge, log=TRUE, prior.count=0.1)
TMM_w <- cpm(x=dge, log=TRUE, prior.count=1.0)
TMM_x <- cpm(x=dge, log=TRUE, prior.count=5.0)
TMM_y <- cpm(x=dge, log=TRUE, prior.count=10.0)
TMM_z <- cpm(x=dge, log=TRUE, prior.count=20.0)
```

## Question 3 plot v
```{r}
qplot(data=as.data.frame(TMM_v), x=Ctl3, y=Ctl5, alpha=I(0.1)) + geom_smooth(method="gam") + geom_abline(color="red")
```

## Question 3 plot w
```{r}
qplot(data=as.data.frame(TMM_w), x=Ctl3, y=Ctl5, alpha=I(0.1)) + geom_smooth(method="gam") + geom_abline(color="red")
```

## Question 3 plot x
```{r}
qplot(data=as.data.frame(TMM_y), x=Ctl3, y=Ctl5, alpha=I(0.1)) + geom_smooth(method="gam") + geom_abline(color="red")
```

## Question 3 plot y
```{r}
qplot(data=as.data.frame(TMM_x), x=Ctl3, y=Ctl5, alpha=I(0.1)) + geom_smooth(method="gam") + geom_abline(color="red")
```

## Question 3 plot z
```{r}
qplot(data=as.data.frame(TMM_z), x=Ctl3, y=Ctl5, alpha=I(0.1)) + geom_smooth(method="gam") + geom_abline(color="red")
```
